---
title: "Using the package lordif to evaluate DIF"
author: "Derek Briggs"
date: "2024-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The document will show you how to evaluate items for DIF using the `lordif` package.

## Load pacakges and data

Once again we will use the CDE data that you used for empirical data analyis #1, but this time we will also include a column that indicates gender of test taker.

```{r data setup}
library(lordif)
setwd("~/Dropbox/Courses/EDUC 8720/Data Sets/MASC-CDE")
fulldata<-read.fwf("CTBMathSci.rwo",widths=c(rep(1,76),3,1,6),skip=2)
cde<-fulldata[,c(1:31,78)]
names(cde)[32]<-"gender"
response<-cde[,1:31]
gender<-as.factor(cde[,32])
#Female = 1, Male = 2 Convert to numeric format for lordif
gender.num<-as.numeric(gender)
```

Before we go any further, always important to look at the frequency distributons of responses by item and group. Items where one group does significantly better than another are ones to keep your eyes on. We ask, could part of this observed difference be caused by bias in the way the item was written?

```{r descriptives}
cross<-list(NULL) #This initializes a new list object
for (i in 1:31){
TAB<-table(cde[,i],cde$gender)
cross[[i]]<-prop.table(TAB,2)
}
cross  
```

## Running the lordif algorithm

```{r lordif}
DIF <- lordif(response, gender.num, criterion = "Chisqr", alpha = 0.01,minCell = 5)
print(DIF)
plot(DIF, labels = c("Female", "Male"),cex=.7)
```

