---
title: "The Verification Frontier"
subtitle: "Where (if anywhere) can we measure critical AI literacy?"
author: "Derek C. Briggs --- SSM 2026 (working simulation)"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4.2, dpi = 130)
```

> **How to read this document.** It is deliberately built to be *audited*, not
> trusted. Critical AI literacy is the ability to catch a plausible-but-wrong
> artifact you cannot verify at a glance --- which is exactly the position you
> are in reading machine-generated code. So every modeling decision below is
> tagged **[CHOICE]**, and Sections 4.2--4.3 *check the math empirically* so you
> are not relying on my algebra being correct. If a claim is load-bearing, it is
> stated as something you can falsify.

# The question

An LLM held at a fixed prompt has its variability set by a dial (temperature);
across the handful of frontier models that variability is being engineered
toward zero. So the durable human variance is not in the model. We conjecture it
lives in **critical AI literacy** (`theta`): the capacity to detect and repair an
output that is wrong but *plausible*.

This document asks a measurement question, not a philosophical one:

> Given a model with capability `c` and "seductiveness" `s`, **how much
> information does a single task carry about a person's `theta`?** Equivalently:
> how many tasks would we need to measure `theta` reliably, and does that number
> grow or shrink as models improve?

# The measurement model

One task = one Bernoulli outcome: *did the final answer come out correct after
the human--AI interaction?* We model the success probability for a person with
critical literacy `theta` as

$$
P(\text{success}\mid\theta) \;=\; \underbrace{c}_{\text{AI right unaided}}
\;+\;(1-c)\,\underbrace{g(\theta)}_{\text{user catches \& fixes a wrong answer}}.
$$

The two terms are the only two ways the final answer succeeds: either the AI was
right on its own (probability `c`), or it fell short (probability `1-c`) and the
user **over-rode** it.

**What "catch" means across domains.** In a *cheaply verifiable* domain (code,
math) "catching" is literal error-detection against an objective check. In a
domain where verification is *expensive or deferred* (the dinner-party story)
there is no such check at hand, so "catch" means **appraisal**: correctly judging
that a fluent, joke-shaped output will actually fall flat for the audience, and
not being captured by its polish. The criterion is then a *deferred, social*
verdict (does the room laugh?) rather than a unit test. So read `g(theta)` as
**the probability the user correctly over-rides a plausible-but-inadequate
answer** --- by detection where a check is cheap, by appraisal where it is not.

The catch-and-repair probability is a standard 2-parameter logistic (IRT) item:

$$
g(\theta) \;=\; \mathrm{logistic}\!\big(a\,(\theta - \delta(s))\big),
\qquad \mathrm{logistic}(x)=\frac{1}{1+e^{-x}} .
$$

- **`theta`** --- the person's critical literacy. **[CHOICE]** population
  distribution `theta ~ N(0,1)`.
- **`a`** --- discrimination of the catch process. **[CHOICE]** `a = 1.5`.
- **`delta(s)`** --- *difficulty of over-riding* a plausible-but-inadequate
  answer, driven by model **seductiveness** `s`: how strongly surface fluency
  masks the (possibly deferred) verdict. More fluent duds are harder to over-ride,
  so `delta` increases in `s`. **[CHOICE]** linear map `s in [0,1] -> delta in [-3, 3]`.
- **`c`** --- model **capability**, the probability the AI is right with no
  human intervention.

> **[CHOICE that you should attack]** Here `theta` is operationalized *narrowly*
> as the ability to over-ride a plausible-but-inadequate answer --- error-detection
> where verification is cheap, appraisal where it is not. Your blog series frames
> critical AI literacy more broadly (calibrated trust, knowing *when* to invoke AI,
> interrogating framing). This model captures one facet. Decide whether that facet
> is the right stand-in for the talk, or whether the construct needs additional
> items of a different kind.

```{r model}
A_DISC <- 1.5                 # [CHOICE] discrimination of the catch process
DELTA_MIN <- -3; DELTA_MAX <- 3   # [CHOICE] seductiveness -> catch-difficulty

logistic <- function(x) 1 / (1 + exp(-x))

# seductiveness s -> difficulty of catching a wrong answer
delta_of_s <- function(s) DELTA_MIN + (DELTA_MAX - DELTA_MIN) * s

# P(final answer correct | theta, c, s)
P_success <- function(theta, c, s) {
  g <- logistic(A_DISC * (theta - delta_of_s(s)))
  c + (1 - c) * g
}
```

## Reading the parameters: an item characteristic curve

Plotting `P(success | theta)` against `theta` for a single item makes `c` and
`delta(s)` concrete. `c` is the **floor** --- the success rate of a person with
no critical literacy, which rises with model capability. `delta(s)` is the
**location**: the critical-literacy level at which success is halfway between the
floor and 1. A more seductive item (higher `s`) has a larger `delta(s)` and
shifts the curve **right** --- harder to over-ride.

```{r icc, fig.width=7, fig.height=4.4}
th <- seq(-4, 4, length.out = 300)
cc <- 0.40                          # lower asymptote (model capability)
s1 <- 0.50; s2 <- 0.75              # two seductiveness levels
d1 <- delta_of_s(s1); d2 <- delta_of_s(s2)   # -> delta = 0 and 1.5
mid <- (cc + 1) / 2                 # halfway between floor and ceiling

plot(th, P_success(th, cc, s1), type = "l", lwd = 3, col = "#1f77b4", ylim = c(0, 1),
     xlab = expression(paste("critical literacy  ", theta)),
     ylab = expression(paste("P(success | ", theta, ")")),
     main = "A hypothetical item characteristic curve")
lines(th, P_success(th, cc, s2), lwd = 3, lty = 2, col = "#d62728")

abline(h = cc, col = "gray55", lty = 3); abline(h = 1, col = "gray80", lty = 3)
text(-3.9, cc - 0.06, paste0("c = ", cc, "  (floor = model capability)"),
     pos = 4, cex = 0.8, col = "gray30")

# delta(s) = theta where the curve passes the halfway point
segments(d1, 0, d1, mid, col = "#1f77b4", lty = 3)
points(d1, mid, pch = 19, col = "#1f77b4"); points(d2, mid, pch = 19, col = "#d62728")
text(d1, mid + 0.05, expression(delta(s)), col = "#1f77b4", pos = 2, cex = 1.0)
arrows(d1 + 0.1, 0.27, d2 - 0.1, 0.27, length = 0.08, col = "#d62728")
text(d2 + 0.1, 0.27, "more seductive", col = "#d62728", pos = 4, cex = 0.78)

legend("bottomright", bty = "n", cex = 0.85, lwd = 3, lty = c(1, 2),
       col = c("#1f77b4", "#d62728"),
       legend = c(paste0("s = ", s1, "  (delta = ", d1, ")"),
                  paste0("s = ", s2, "  (delta = ", d2, ")")))
```

## Two domains: where capability can and cannot rise {#domains}

The single most important refinement: **`c` does not mean the same thing in
every domain.** The axis that matters is not a clean verifiable/unverifiable
binary but the **cost and latency of verification** --- how cheaply, quickly, and
objectively the quality of an output can be checked. Write `v in [0,1]` for the
*cheapness of verification*:

- **Cheaply verifiable** (`v = 1`: code compiles, the theorem checks, the
  arithmetic ties out). An objective signal is available at once, so training can
  push capability close to 1. As `c -> ~0.96` the opportunity term `(1-c) -> 0`,
  and by Check 3 the information about `theta` collapses. Critical literacy decays
  to "did you bother to run the test."
- **Expensively / deferredly verifiable** (`v = 0`: is this joke funny for *my*
  dinner party? is this framing wise? is this prose any good?). A verifier still
  exists --- the room laughs or it doesn't --- but it is **deferred, noisy, and
  audience-specific**. Two consequences: (i) training cannot cleanly optimize
  against so sparse and after-the-fact a signal, so capability **plateaus below
  1** at `c_cap` --- the model floor-raises *fluency*, not *quality*; and (ii)
  that same deferred verdict is the only scoring key for the human's judgment, so
  critical literacy here is **appraisal** (seeing past fluency to anticipate the
  verdict) and remains the main thing between the user and a confident,
  plausible dud.

(*Genuinely* unverifiable output --- pure preference with no convergent verdict
even in principle --- would have no scoring key at all, so critical literacy there
would be **unmeasurable**, not merely expensive to measure. The interesting,
measurable zone is real-quality-but-expensively-verified.)

We trace one model-progress path per domain. Generation `g in [0,1]` runs early
-> frontier. Capability rises to a domain-dependent ceiling; seductiveness rises
with fluency, **shared** across domains (a deliberate simplification --- see
caveats). The two roads start at the *same* early model and fork only because
cheaply-verifiable capability can keep climbing while expensively-verifiable
capability cannot.

```{r domains}
C0     <- 0.50   # [CHOICE] early capability, both domains
C_CEIL <- 0.96   # [CHOICE] verifiable capability ceiling (objective signal -> ~1)
C_CAP  <- 0.55   # [CHOICE] unverifiable capability ceiling (no signal -> plateau)
S0     <- 0.45   # [CHOICE] early seductiveness
S_MAX  <- 0.62   # [CHOICE] frontier seductiveness (fluency rises with scale)

cmax_of_v <- function(v) C_CAP + (C_CEIL - C_CAP) * v   # capability ceiling by domain
c_traj <- function(g, v) C0 + (cmax_of_v(v) - C0) * g   # capability over generations
s_traj <- function(g)    S0 + (S_MAX - S0) * g          # seductiveness over generations
```

# Why this model encodes the thesis (claims to verify)

These are predictions you can check against the code, not assertions to accept:

1. **Capability erases the construct.** As `c -> 1`, `P -> 1` for *every*
   `theta`, so the curve flattens and information about `theta` must go to 0.
   A strong model on a checkable task leaves nothing to measure.
2. **A verification frontier exists.** For a population centered at `theta = 0`,
   a task is most informative when the catch-difficulty is matched to the
   population, `delta(s) ~ 0`, i.e. at an *intermediate* seductiveness. Too-easy
   errors (everyone catches) and too-hard errors (nobody catches) are both
   uninformative.

# Fisher information: the per-task "measurement value"

For a single Bernoulli item with success probability `P(theta)`, the Fisher
information a response carries about `theta` is

$$
I(\theta) \;=\; \frac{\big(P'(\theta)\big)^2}{P(\theta)\,\big(1-P(\theta)\big)},
\qquad P'(\theta)=\frac{dP}{d\theta}.
$$

Differentiating our `P`:

$$
P'(\theta) = (1-c)\,a\,g(\theta)\big(1-g(\theta)\big).
$$

We report the **population-expected information**
$\mathrm{E}_\theta[\,I(\theta)\,]$ over `theta ~ N(0,1)`: how much one task
tells us about a randomly drawn person --- i.e. the per-item value of the task
as a measurement instrument.

```{r info-functions}
# grid quadrature for theta ~ N(0,1)
THETA <- seq(-4, 4, length.out = 401)
W <- dnorm(THETA); W <- W / sum(W)         # normalized N(0,1) weights

# analytic dP/dtheta
dP_dtheta <- function(theta, c, s) {
  g <- logistic(A_DISC * (theta - delta_of_s(s)))
  (1 - c) * A_DISC * g * (1 - g)
}

# Fisher information at a single theta
info_point <- function(theta, c, s) {
  P  <- P_success(theta, c, s)
  dP <- dP_dtheta(theta, c, s)
  denom <- pmax(P * (1 - P), 1e-9)
  dP^2 / denom
}

# population-expected information for one task
expected_information <- function(c, s) sum(W * info_point(THETA, c, s))
```

## Check 1 --- is `dP/dtheta` actually the derivative of `P`?

Before trusting the information formula, verify the analytic derivative against a
finite-difference approximation. If my calculus is wrong, this fails loudly.

```{r check-derivative}
set.seed(1)
eps <- 1e-6
ok <- TRUE
for (i in 1:200) {
  th <- runif(1, -3, 3); c <- runif(1, .1, .95); s <- runif(1, .05, .95)
  num <- (P_success(th + eps, c, s) - P_success(th - eps, c, s)) / (2 * eps)
  ana <- dP_dtheta(th, c, s)
  if (abs(num - ana) > 1e-5) ok <- FALSE
}
stopifnot(ok)
cat("PASS: analytic dP/dtheta matches finite differences.\n")
```

## Check 2 --- is the Fisher-information *formula* right?

The Fisher information equals the variance of the score
$\partial_\theta \log P(Y\mid\theta)$. For a Bernoulli outcome the score is
$(Y-P)\,P' / [P(1-P)]$. We simulate outcomes at a fixed `theta` and confirm the
empirical score variance matches `info_point()`.

```{r check-fisher}
mc_information <- function(theta, c, s, n = 2e6) {
  P  <- P_success(theta, c, s)
  dP <- dP_dtheta(theta, c, s)
  Y  <- rbinom(n, 1, P)
  score <- (Y - P) * dP / (P * (1 - P))   # d/dtheta log-likelihood
  var(score)
}
set.seed(7)
for (case in list(c(0.3, 0.5, 0.4), c(0.6, 0.9, 0.55), c(-0.5, 0.7, 0.3))) {
  th <- case[1]; c <- case[2]; s <- case[3]
  cat(sprintf("theta=%+.1f c=%.2f s=%.2f  analytic=%.4f  monte-carlo=%.4f\n",
              th, c, s, info_point(th, c, s), mc_information(th, c, s)))
}
```

The Monte-Carlo column should track the analytic column to ~2 decimals. If it
does, the per-task information function is trustworthy.

## Check 3 --- does information collapse as `c -> 1`? (Claim 1)

```{r check-collapse}
sapply(c(0.3, 0.6, 0.9, 0.99, 0.999),
       function(c) expected_information(c, s = 0.5))
```

These should march monotonically toward 0 --- confirming that a more capable
model on a fixed task carries *less* information about critical literacy.

# Panel A: where the trait is measurable

```{r surface, fig.width=6.5, fig.height=5}
cs <- seq(0.05, 0.98, length.out = 120)
ss <- seq(0.02, 0.98, length.out = 120)
INFO <- outer(cs, ss, Vectorize(expected_information))   # rows=c, cols=s

# verification-frontier ridge: best s for each c
ridge_s <- ss[apply(INFO, 1, which.max)]

# two roads across the same map, forked by verifiability
GEN <- seq(0, 1, length.out = 6)                 # early -> frontier
ver_c <- c_traj(GEN, 1); ver_s <- s_traj(GEN)    # verifiable domain (v=1)
unv_c <- c_traj(GEN, 0); unv_s <- s_traj(GEN)    # unverifiable domain (v=0)

filled.contour(
  x = cs, y = ss, z = INFO, color.palette = function(n) hcl.colors(n, "viridis"),
  xlab = "model capability  c  (P right unaided)",
  ylab = "model seductiveness  s  (subtlety of errors)",
  main = "A. Two roads across the same map",
  key.title = title(main = "info", cex.main = 0.8),
  plot.axes = {
    axis(1); axis(2)
    lines(cs, ridge_s, col = "white", lwd = 2, lty = 2)           # verification frontier
    lines(ver_c, ver_s, col = "white", lwd = 2)
    points(ver_c, ver_s, pch = 21, bg = "#1f77b4", col = "white", cex = 1.1)
    lines(unv_c, unv_s, col = "white", lwd = 2)
    points(unv_c, unv_s, pch = 21, bg = "#d62728", col = "white", cex = 1.1)
    # mark the shared start and the two frontiers (short, interior labels)
    points(ver_c[1], ver_s[1], pch = 21, bg = "white", col = "black", cex = 1.3)
    legend("topleft", bty = "n", text.col = "white", cex = 0.8,
           legend = c("verification frontier (ridge)",
                      "verifiable road (stats / code)",
                      "unverifiable road (the joke)",
                      "shared early model"),
           lty = c(2, 1, 1, NA), pch = c(NA, 21, 21, 21), lwd = c(2, 2, 2, NA),
           col = "white", pt.bg = c(NA, "#1f77b4", "#d62728", "white"))
  })
```

White dashed line = the verification frontier (most-informative `s` per `c`);
white solid line with red points = the model-progress path.

# Panel B: the window narrows for verifiable tasks, not unverifiable ones

The info-vs-`s` curve is a function of `c` and `s` only --- the surface is the
same map for any domain. What `v` controls is *which* `c` a domain can reach. So
we slice the map at the three capabilities the two roads actually occupy: the
shared early model, the unverifiable frontier (`C_CAP`), and the verifiable
frontier (`C_CEIL`).

```{r panelB}
infoB_early <- sapply(ss, function(s) expected_information(C0,     s))  # 0.50, both domains, early
infoB_unv   <- sapply(ss, function(s) expected_information(C_CAP,  s))  # 0.55, unverifiable frontier
infoB_ver   <- sapply(ss, function(s) expected_information(C_CEIL, s))  # 0.96, verifiable frontier

plot(ss, infoB_early, type = "l", lwd = 2.4, col = "#444444",
     ylim = c(0, max(infoB_early)),
     xlab = "model seductiveness  s",
     ylab = "info about critical literacy (per task)",
     main = "B. The window narrows for verifiable tasks, not unverifiable ones")
lines(ss, infoB_unv, lwd = 2.8, col = "#d62728")
lines(ss, infoB_ver, lwd = 2.8, col = "#1f77b4")
legend("topright", bty = "n", cex = 1.0, lwd = 2.4,
       legend = c("c = 0.50", "c = 0.55", "c = 0.96"),
       col = c("#444444", "#d62728", "#1f77b4"))
```

The unverifiable frontier (red) sits almost on top of the early model (grey): in
deferredly-verified domains the measurement window barely narrows. Only the
verifiable frontier (blue) collapses --- the same `v`-contrast as Panel A, now
sliced along `s`.

# Panel C: is AI literacy a transient construct?

Translate information into a tangible cost. For a trait with variance 1, total
information `I_tot` gives reliability `rho = I_tot / (I_tot + 1)`, so reaching
`rho = 0.80` needs `I_tot = 4`. The number of trap-items required is
`4 / E[info per item]`.

```{r panelC}
RHO <- 0.80
I_NEED <- RHO / (1 - RHO)                     # = 4
items_for_reliability <- function(c, s) {
  e <- expected_information(c, s)
  if (e <= 1e-6) Inf else I_NEED / e
}

ver_info <- mapply(expected_information, ver_c, ver_s)   # verifiable road
unv_info <- mapply(expected_information, unv_c, unv_s)   # unverifiable road
gens <- seq_along(GEN)

plot(gens, unv_info, type = "b", pch = 19, col = "#d62728", lwd = 2.3,
     ylim = c(0, max(unv_info, ver_info)),
     xlab = "model generation  (early -> frontier)",
     ylab = "info per task about critical literacy",
     main = "C. Verifiable literacy decays; unverifiable persists")
lines(gens, ver_info, type = "b", pch = 17, col = "#1f77b4", lwd = 2.3)
legend("right", c("unverifiable (the joke)", "verifiable (stats / code)"),
       col = c("#d62728", "#1f77b4"), pch = c(19, 17), lwd = 2.3, bty = "n", cex = 0.85)
```

## The headline table

```{r table}
tab <- data.frame(
  point = c("shared early model", "verifiable frontier (stats/code)",
            "unverifiable frontier (the joke)"),
  c = c(ver_c[1], ver_c[6], unv_c[6]),
  s = c(ver_s[1], ver_s[6], unv_s[6]),
  info_per_item = c(expected_information(ver_c[1], ver_s[1]),
                    expected_information(ver_c[6], ver_s[6]),
                    expected_information(unv_c[6], unv_s[6])),
  items_for_rho_.80 = c(items_for_reliability(ver_c[1], ver_s[1]),
                        items_for_reliability(ver_c[6], ver_s[6]),
                        items_for_reliability(unv_c[6], unv_s[6]))
)
knitr::kable(tab, digits = c(0, 2, 2, 4, 0))

prog <- data.frame(generation = gens,
                   ver_c = ver_c, ver_info = ver_info,
                   unv_c = unv_c, unv_info = unv_info,
                   ver_items = round(mapply(items_for_reliability, ver_c, ver_s)),
                   unv_items = round(mapply(items_for_reliability, unv_c, unv_s)))
knitr::kable(prog, digits = c(0, 2, 4, 2, 4, 0, 0))
```

# An item-bank version: seductiveness vs. item difficulty {#bank}

The model so far uses a *single* over-ride difficulty per generation,
`delta_of_s(s)` --- a one-item caricature. A real design has a **bank** of
prompts whose errors vary in how seductive they are. So we **promote
`delta_of_s(s)` from "the difficulty" to the _mean_ of an item-difficulty
distribution**, and draw a bank around it:

$$
\delta_i \sim \mathrm{N}\big(\delta(s),\; \sigma_\delta\big),
\qquad i = 1, \dots, N_\mathrm{items} .
$$

Capability `c` stays fixed per generation (your call this round). A key
consequence: because seductiveness `s` is shared across the two domains, **they
draw from the _same_ difficulty bank** and differ only in the model that scores
it (`c = C_CAP` vs `c = C_CEIL`).

```{r bank-setup}
SIGMA_DELTA <- 1.0     # [CHOICE] spread of item difficulties (logits) around the mean
N_ITEMS     <- 200     # [CHOICE] prompts in the bank
BANK_SEED   <- 2026

# per-item info as a function of difficulty delta DIRECTLY (theta ~ N(0,1))
info_point_delta <- function(theta, c, delta) {
  g  <- logistic(A_DISC * (theta - delta))
  P  <- c + (1 - c) * g
  dP <- (1 - c) * A_DISC * g * (1 - g)
  dP^2 / pmax(P * (1 - P), 1e-9)
}
ei_delta <- function(c, delta) sum(W * info_point_delta(THETA, c, delta))

# consistency: the representative-item model is just the single-item case
stopifnot(abs(ei_delta(0.55, delta_of_s(0.62)) -
              expected_information(0.55, 0.62)) < 1e-9)
cat("PASS: ei_delta reproduces the representative-item information.\n")
```

```{r bank-functions}
# exact bank expectation: average per-item info over delta ~ N(delta_of_s(s), sigma)
DELTA_GRID <- seq(-7, 7, length.out = 281)
bank_info_exact <- function(c, s, sigma = SIGMA_DELTA) {
  wd <- dnorm(DELTA_GRID, delta_of_s(s), sigma); wd <- wd / sum(wd)
  sum(wd * sapply(DELTA_GRID, function(d) ei_delta(c, d)))
}

# a concrete RANDOM bank of N_ITEMS prompts
draw_bank <- function(c, s, sigma = SIGMA_DELTA, n = N_ITEMS, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  delta_i <- rnorm(n, delta_of_s(s), sigma)
  list(delta = delta_i,
       info  = sapply(delta_i, function(d) ei_delta(c, d)))
}
```

## Same bank, two capabilities

The same trap-items are highly informative scored through a weak model and
nearly useless scored through a strong one --- the divergence is entirely in the
lower asymptote, not the difficulties.

```{r bank-illustrate, fig.width=7, fig.height=3.4}
s_front <- S_MAX                              # shared frontier seductiveness (= 0.62)
set.seed(BANK_SEED)
delta_i  <- rnorm(N_ITEMS, delta_of_s(s_front), SIGMA_DELTA)
info_unv <- sapply(delta_i, function(d) ei_delta(C_CAP,  d))   # c = 0.55 (unverifiable)
info_ver <- sapply(delta_i, function(d) ei_delta(C_CEIL, d))   # c = 0.96 (verifiable)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
hist(delta_i, breaks = 20, col = "gray70", border = "white",
     main = "Shared item bank", xlab = "delta_i  (logits; 0 = population mean)")
abline(v = 0, lty = 2)
boxplot(list(unverifiable = info_unv, verifiable = info_ver),
        col = c("#d62728", "#1f77b4"), ylab = "per-item info about theta",
        main = "Same bank, two models")
par(mfrow = c(1, 1))

cat(sprintf("random bank  : unverifiable mean=%.4f (items=%.0f) | verifiable mean=%.4f (items=%.0f)\n",
            mean(info_unv), 4/mean(info_unv), mean(info_ver), 4/mean(info_ver)))
cat(sprintf("exact bank   : unverifiable     =%.4f             | verifiable     =%.4f\n",
            bank_info_exact(C_CAP, s_front), bank_info_exact(C_CEIL, s_front)))
```

## The divergence survives a real bank

```{r bank-panelC}
ver_info_bank <- sapply(seq_along(GEN), function(i) bank_info_exact(ver_c[i], ver_s[i]))
unv_info_bank <- sapply(seq_along(GEN), function(i) bank_info_exact(unv_c[i], unv_s[i]))

plot(gens, unv_info_bank, type = "b", pch = 19, col = "#d62728", lwd = 2.3,
     ylim = c(0, max(unv_info_bank, unv_info)),
     xlab = "model generation  (early -> frontier)",
     ylab = "bank-average info per task",
     main = "C'. Item-bank version: the divergence survives")
lines(gens, ver_info_bank, type = "b", pch = 17, col = "#1f77b4", lwd = 2.3)
lines(gens, unv_info, lty = 3, col = "#d62728")     # representative-item, for comparison
lines(gens, ver_info, lty = 3, col = "#1f77b4")
legend("topright", bty = "n", cex = 0.8,
       legend = c("unverifiable (bank)", "verifiable (bank)", "representative item (dotted)"),
       col = c("#d62728", "#1f77b4", "gray40"), pch = c(19, 17, NA),
       lty = c(1, 1, 3), lwd = 2)
```

Bank-average information is **lower** than the representative-item value (the
bank includes off-target prompts), so items-needed rises modestly --- the joke
goes from ~50 to ~55, the verifiable frontier from ~719 to ~769 --- but the
verifiable-vs-unverifiable gap is unchanged.

## Sampling variability of a random bank

Because the bank is drawn at random, the items-needed estimate has its own
uncertainty. This is honest measurement bookkeeping --- and a reminder that
"50 items" is itself an estimate.

```{r bank-variability}
set.seed(BANK_SEED)
reps <- replicate(300, {
  d <- rnorm(N_ITEMS, delta_of_s(s_front), SIGMA_DELTA)
  4 / mean(sapply(d, function(x) ei_delta(C_CAP, x)))
})
cat(sprintf("Unverifiable frontier, items-needed across 300 random banks of %d prompts:\n  median=%.0f   90%% interval [%.0f, %.0f]\n",
            N_ITEMS, median(reps), quantile(reps, .05), quantile(reps, .95)))
```

Wider `SIGMA_DELTA` scatters prompts further from the informative region,
lowering average information and raising items-needed: that is the design lever
for *how tightly the prompt bank should target the population.* The "moving
frontier" becomes concrete here --- as `s` drifts up across generations, you
must re-center `delta_of_s(s)` and re-cover the population, or the bank decays.

# Honest caveats (the places to push)

1. **The verifiable/unverifiable split is now *in* the model** (Section
   \@ref(domains)), via a capability ceiling `c_cap` for unverifiable domains.
   The remaining simplification is that seductiveness `s` is *shared* across
   domains. A fuller model would also let verification *suppress effective
   seductiveness* in verifiable domains (you can catch any error if you run the
   check), which would push the verifiable road off the ridge even faster. The
   capability ceiling alone already produces the divergence, so I kept `s` shared
   to isolate one mechanism. **Attack the ceilings:** `C_CAP = 0.55` is a guess --- change it
   in the Section \@ref(domains) chunk and re-knit to move the unverifiable
   plateau; the `sensitivity()` block below varies `a` and the seductiveness range.
2. **`delta(s)` linear, `a` constant, `theta ~ N(0,1)`** are all conveniences.
   Re-run the sliders below to see whether conclusions survive.
3. **Independence and unidimensionality.** Items are treated as conditionally
   independent given a single `theta`. If catching errors is multidimensional
   (domain priors vs. tool fluency --- the Chollet confound), this overstates
   information.
4. **`theta` = over-riding a plausible-but-inadequate answer** (detection where
   verification is cheap, appraisal where it is deferred). See the construct
   caveat in Section 2. Note the scoring-key dependence: measuring `theta` in an
   expensively-verified domain needs the *deferred* verdict (e.g. the audience)
   as a criterion --- so "50 items" understates the real cost, and a domain with
   no convergent verdict at all would make `theta` unmeasurable, not just costly.

```{r sliders}
# Re-run the whole story under your own assumptions.
# NOTE: this temporarily sets the global parameters, then RESTORES them on exit,
# so calling it leaves no side effects (an earlier version leaked global state --
# see the lesson in the prose above).
sensitivity <- function(a = 1.5, dmin = -3, dmax = 3, rho = 0.80) {
  old <- list(A_DISC, DELTA_MIN, DELTA_MAX)
  on.exit({ A_DISC <<- old[[1]]; DELTA_MIN <<- old[[2]]; DELTA_MAX <<- old[[3]] })
  A_DISC <<- a; DELTA_MIN <<- dmin; DELTA_MAX <<- dmax
  data.frame(
    verifiable_frontier = (rho/(1-rho)) / expected_information(0.96, 0.62),
    unverifiable_frontier = (rho/(1-rho)) / expected_information(0.55, 0.62),
    shared_early = (rho/(1-rho)) / expected_information(0.50, 0.45)
  )
}
sensitivity()                       # defaults (delta -3..3, a=1.5)
sensitivity(a = 1.0)                # less discriminating catch process
sensitivity(dmin = -2, dmax = 2)    # narrower seductiveness range
```

# What to take to the talk

- **Two roads, one map.** The same model progress that drives critical literacy
  *off* the measurable ridge in verifiable domains (code/stats) keeps it *on* the
  ridge in unverifiable ones (the joke) --- because capability can climb toward 1
  only where there is a ground-truth signal to climb. See the headline table for
  the trap-items each road needs.
- **The construct is most measurable exactly where the model can help least.**
  Critical AI literacy is durable in domains where the AI cannot verify itself ---
  judgment, taste, framing, wisdom --- which is both why it matters and why it is
  hard to certify.
- **A live empirical question, not a settled one.** Whether unverifiable-domain
  literacy truly persists, or eventually erodes as fluency outruns human
  detection (`s` past the catchable band), is the design that a future study
  would test. That is Chollet's generalization frontier relocated onto the human.
